Regression Cube Analysis of Cohort Study Data

### 2015-06-20 draft.tex
covered orderings as well as advantages/disadvanteges of the mini map

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al.~\cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates as well as time plots to display and brush the data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
%%We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.
%%
Generalized Pairs plots (\texttt{GPLOM'S}) extend the idea of scatter plot matrices by the pairwise depiction of heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter.
%%
They are therefore suitable for \emph{small} epidemiological data sets.
%%
Dai et al.~\cite{Dai2005} incorporate a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with \emph{Pearsson's} $r$ values.
%%
A \emph{concept map} summarizes features related to a specified disease.
%%
Time-dependent epidemiological data are visualized by Chui et al.~\cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender with regard to influenza- and salmonellosis-associated hospitalizations.
%%
%We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%
\paragraph{Statistical Analysis.}
%%
%Statistical analysis is an inherent part of Visual Analysis.
%%
%Here we present work showing a strong focus on statistical analysis in this context.
%% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al.~\cite{Bertini} present an overview of quality metrics describing high dimensional-data.
%%
Their research agenda acts as guideline for our design.
%%
Namely, comprising perceptual tuning with human pattern recognition of important aspects, scalability between different data set sizes and application testing with domain experts.
%% The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems
Ahmadi et al.~\cite{Ahmadi} define the \emph{Sparse Regression Cube} that partitions sparse high dimensional data into subspaces, which are then described by their most reliable linear regression model.
%%
%In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on an algebraic representation for efficient regression-model calculation to find the best fit for a subspace.
%%
%The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

@@

\paragraph{3D Prism as Data Mini-Map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular with our domain experts, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The 3D cube representation \design{was redesigned to} act as an overview over the whole data set. 
%%
Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D mosaic plot.
%%
\design{The distinction between overview and details and demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its $R^2$ features and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\paragraph{Tackling the disadvantages of 3D information visualization.}
%%
3D information visualizations are often criticized for introducing occlusions and interaction problems.
%%
These are often not balanced out by the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these problems.
%%
Since the $R^2$ values are mapped on data point opacity, large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low $R^2$ values.
%%
Also, the preceding correlation-based feature selection reduces the information space significantly, leading to sparse cubes.
%%
Overlapping is still an issue, but this way greatly reduced in its effect to the readability of the visualization.
%%

@@

%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set analysis.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientists responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present the last study wave.

>>>>

%% Steenwijk: Integrated Visual Analysis for Heterogeneous Datasets in Cohort Studies
Steenwijk et al.~\cite{Steenwijk} focus on \emph{hypothesis-free} exploration of cohort data.
%%
They employ a framework consisting of feature extraction and visualization to derive dependencies between image- and non-image features.
%%
They incorporate linked views using scatter plots, bar charts, parallel coordinates as well as time plots to display and brush the data.
%%
%Medical image data can be incorporated to investigate potentially interesting individuals.
%%
%%We distinguish to this and other linked-view analysis based approaches by showing abstracted information derived from a preceding analyze-first step to highlight relationships rather than displaying the actual data using various information visualization techniques.
%%
Generalized Pairs plots (\texttt{GPLOM'S}) extend the idea of scatter plot matrices by the pairwise depiction of heterogeneous data using type-combination-dependent visualizations \cite{GPLOMS, Francois2013}.
%%
The techniques are useful for selected features or a data set with a small number of features due to their increase in complexity with every additional parameter.
%%
They are therefore suitable for \emph{small} epidemiological data sets.
%%
Dai et al.~\cite{Dai2005} incorporate a \texttt{GPLOM}-like visualization using choropleth maps mapping spatial data, such as mortality rates together with scatter plots augmented with \emph{Pearsson's} $r$ values.
%%
A \emph{concept map} summarizes features related to a specified disease.
%%
Time-dependent epidemiological data are visualized by Chui et al.~\cite{Chui2011} using multi-panel graphs highlighting risk factor differences with age and gender with regard to influenza- and salmonellosis-associated hospitalizations.
%%
%We also include confounder features (age, gender) into the analysis while focusing on non-time-dependent cohort study data sets.
%%
\paragraph{Statistical Analysis.}
%%
%Statistical analysis is an inherent part of Visual Analysis.
%%
%Here we present work showing a strong focus on statistical analysis in this context.
%% Quality Metrics in High-Dimensional Data Visualization - An Overview and Systematization
Bertini et al.~\cite{Bertini} present an overview of quality metrics describing high dimensional-data.
%%
Their research agenda acts as a guideline for our design.
%%
Namely, comprising perceptual tuning with human pattern recognition of important aspects, scalability between different data set sizes and application testing with domain experts.
%% The Sparse Regression Cube: A Reliable Modeling Technique for Open Cyber-physical Systems
Ahmadi et al.~\cite{Ahmadi} define the \emph{Sparse Regression Cube} that partitions sparse high dimensional data into subspaces, which are then described by their most reliable linear regression model.
%%
%In their application, they try to find most fuel-efficient roads connecting two user-defined landmarks.
%%
They focus on an algebraic representation for efficient regression-model calculation to find the best fit for a subspace.
%%
%The dimensions for the calculations are fixed, while we assess different features for each model calculation and can therefore not benefit from a hierarchical calculation approach.

@@

\paragraph{3D Prism as Data Mini-Map.}
\design{
%%
In early prototype versions, the 3D prism acted as starting point for the data analysis without the implementation of a separate 2D view.
%%
Slices were shown using cutaway planes.
%%
This approach was not popular with our domain experts, because the complexity of the visualization overwhelmed them. %did not like the combination of 3D interaction and slicing with representatives.
%%
}
%%
The 3D cube representation \design{was redesigned to} act as an overview over the whole data set. 
%%
Its purpose is not to derive detailed information about data points.
%%
It serves as a function similar to a mini-map, guiding the attention to points of interest in the data.
%%
It also gives context information about adjacent data values when using the 2D mosaic plot.
%%
\design{The distinction between overview and details and demand using two different representations was well received with our domain experts.}
%%
The displayed prism shows values above the matrix diagonal.
%%
For formulas with a dynamic target feature (e.g. exploratory analysis using $Z \sim X + Y$), the color encodes the absolute regression metric values (Fig.~\ref{fig:System}b).
%%
Applying this strategy to a formula containing a static target (e.g. $Cancer \sim X + Y + Z$) yields many occlusions, since the CFS algorithm creates the same feature space for every slice.
%%
For such formulas, the 3D view encodes every data element as absolute difference between its $R^2$ features and the global mean along the z-axis.
%%
This highlights slices with unusually low or high results (Fig.~\ref{fig:Application}).
%%
\magenta{Variables are ordered the same way in the 2D mosaic plot and 3D cube to preserve the mental model and make them visually analogous.}

\paragraph{Tackling the disadvantages of 3D information visualization.}
%%
3D information visualizations are often criticized for introducing occlusions and interaction problems.
%%
These are often not balanced out by the advantages of using the third dimension for visual mapping.
%%
We aim to minimize these problems.
%%
Since the $R^2$ values are mapped on data point opacity, large values are highlighted in the prism, guiding the focus to the respective slices.
%%
It also creates a sparse representation, since the majority of regression models yield (depending on the data set and the chosen formula) low $R^2$ values.
%%
Also, the preceding correlation-based feature selection reduces the information space significantly, leading to sparse cubes.
%%
Overlapping is still an issue, but this way greatly reduced in its effect to the readability of the visualization.
%%

@@

%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast fat data set analysis.
%%
She is a radiologist, responsible for the SHIP-MRI acquisition and also responsible of the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience) assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants for which the MRI data was available to that point.
%%
The MRI scans for each subject have been introduced in SHIP-2, so the target feature is only present the last study wave.
